---
title: "cBioPortal_test"
author: "Xingjun Liu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description

This is an R Markdown for test pulling gene expression data through [cBioPortal API](https://waldronlab.io/cBioPortalData/reference/cBioPortal.html).

The script include:

1. finds a TCGA glioma study,

2. locates the mutation profile and the mRNA expression profile,

3. pulls mutation data to identify IDH1-R132H mutant samples,

4. pulls expression (z-scores if available) for the samples,

5. computes mean expression difference (mutant − wild-type),

6. returns & plots the top 10 overexpressed genes.

This uses the exact **cBioPortalData** functions documented in the reference *(cBioPortal(), getStudies(), molecularProfiles(), geneTable(), mutationData(), getSampleInfo(), getDataByGenes() / fetchData() as appropriate)*

## Install and load required libraries
```{r library, echo=TRUE,eval=TRUE}

# Install & load necessary packages
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("cBioPortalData")

library(cBioPortalData)
library(tidyr)
library(ggplot2)
library(dplyr)

```
## 1. Build API connection and pull all studies 
```{r connection, echo = TRUE, eval=TRUE}
# Initialize connection to the API
cbio <- cBioPortal()

# List all studies, then filter for glioma/TCGA
studies <- getStudies(cbio, buildReport = FALSE)

```
## 2. Pull glioma studies include: glioma, lgg, gbm from Cbioportal stuies
```{r glioma_stuies, echo=TRUE,eval=TRUE}
glioma_candidates <- studies %>%
  filter(grepl("glioma", name, ignore.case = TRUE) |
           grepl("lgg", studyId, ignore.case = TRUE) |
           grepl("gbm", studyId, ignore.case = TRUE) |
           grepl("glioma", studyId, ignore.case = TRUE))

print(head(glioma_candidates[, c("studyId", "name")], 20))

# choose study (change if you prefer "gbm_tcga" or combine later)
study_id <- "lgg_tcga"   # <- change if you want another

```
## 3. Find mutation & expression molecular profile IDs for the study

```{r expr_profile, echo=TRUE,eval=TRUE}
# mutation & expression molecular profile
profiles <- molecularProfiles(cbio, studyId = study_id)
# prefer mutation profile (MAF / MUTATION_EXTENDED)
mut_profiles <- profiles %>% filter(grepl("MUTATION", molecularAlterationType, ignore.case = TRUE) |
                                      grepl("mutations", molecularProfileId, ignore.case = TRUE) |
                                      grepl("maf", datatype, ignore.case = TRUE))
if (nrow(mut_profiles) == 0) stop("No mutation profile found for study: ", study_id)
mutation_profile_id <- mut_profiles$molecularProfileId[1]
message("Using mutation profile: ", mutation_profile_id)

# prefer mRNA z-score profile (for comparison)
expr_profiles <- profiles %>% filter(grepl("MRNA_EXPRESSION", molecularAlterationType, ignore.case = TRUE))
# try to pick a Z-SCORE profile if present (cBioPortal's Comparison tab uses z-scores)
expr_profile_id <- expr_profiles %>%
  filter(grepl("Z-SCORE", datatype, ignore.case = TRUE)) %>%
  pull(molecularProfileId) %>% .[1]
if (is.na(expr_profile_id) || length(expr_profile_id) == 0) {
  expr_profile_id <- expr_profiles$molecularProfileId[1]
}
message("Using expression profile: ", expr_profile_id)

```
## 4. Get Entrez ID for IDH1,gene panel and sample information
```{r genes, echo=TRUE,eval=TRUE}
# ---get Entrez ID for IDH1 ----
gene_table <- geneTable(cbio, pageSize = 20000)   # pull gene table
idh1_entrez <- gene_table$entrezGeneId[match("IDH1", gene_table$hugoGeneSymbol)]
if (is.na(idh1_entrez)) stop("IDH1 not found in gene table on this cBioPortal instance")

# Collect sample information
sample_info <- getSampleInfo(cbio, studyId = study_id)

# Extract just the sample IDs
sample_ids <- sample_info$sampleId

```
## 5. Get all mutation data for IDH1 and extract sample information
```{r mutation-data, echo=TRUE,eval=TRUE}
# ----query mutation data for IDH1 and extract samples with R132H ----
mut_df <- mutationData(cbio,
                       molecularProfileIds = mutation_profile_id,
                       entrezGeneIds = idh1_entrez,
                       sampleIds = sample_ids)

# look for 'R132H' in protein-change column (handle variants like 'p.R132H')
mutant_samples <- mut_df[["lgg_tcga_mutations"]][["sampleId"]]
message("Found ", length(mutant_samples), " samples with IDH1 R132H")

if (length(mutant_samples) == 0) {
  stop("No IDH1 R132H mutants found. You may want to broaden the mutation filter or inspect mut_df manually.")
}

```

## 6. Get all samples for study and define wild-type group, fetch expression data for genes

```{r echo=TRUE,eval=TRUE}
# ---- get all samples for study and define wild-type group ----
# sample_info <- getSampleInfo(cbio, studyId = study_id)
all_samples <- sample_info$sampleId
wt_samples <- setdiff(all_samples, mutant_samples)
message("Total study samples: ", length(all_samples), "; wild-type candidates: ", length(wt_samples))

# ----  fetch expression data for genes (we'll fetch all genes then compute ranking) ----
# NOTE: getDataByGenes allows using Hugo symbols (faster than entrez in many cases)
all_genes <- gene_table$hugoGeneSymbol

# Query expression (may take a bit of time for all genes)
expr_raw <- getDataByGenes(cbio,
                           studyId = study_id,
                           genes = all_genes,
                           by = "hugoGeneSymbol",
                           molecularProfileIds = expr_profile_id,
                           sampleIds = c(mutant_samples, wt_samples))

# expr_raw should be a data.frame/tibble with columns for gene, sampleId, and value (z-score or expression)

if (is.list(expr_raw) && length(expr_raw) > 0 && is.data.frame(expr_raw[[1]])) {
  expr_df <- expr_raw[[1]]
} else if (is.data.frame(expr_raw)) {
  expr_df <- expr_raw
} else {
  stop("Unexpected structure returned by getDataByGenes(); inspect the object 'expr_raw'.")
}

# locate columns and rename
gene_col_expr <- names(expr_df)[grepl("hugoGene", names(expr_df), ignore.case = TRUE)][1]
sample_col_expr <- names(expr_df)[grepl("sampleId", names(expr_df), ignore.case = TRUE)][1]
value_col_expr <- names(expr_df)[grepl("value|zscore|z_score|z-score|score", names(expr_df), ignore.case = TRUE)][1]

if (any(is.na(c(gene_col_expr, sample_col_expr, value_col_expr)))) {
  stop("Can't find gene/sample/value columns in expression result; names(expr_df): ", paste(names(expr_df), collapse = ", "))
}

# convert to matrix: genes x samples
expr_df2 <- expr_df %>%
  select(gene = all_of(gene_col_expr),
         sampleId = all_of(sample_col_expr),
         value = all_of(value_col_expr)) %>%
  filter(!is.na(value))

# use tapply to reshape to matrix (handles duplicates by taking the first)
expr_mat <- tapply(expr_df2$value, list(expr_df2$gene, expr_df2$sampleId), function(x) x[1])
expr_mat <- as.matrix(expr_mat)

# keep only samples that appear in expression matrix
mut_samples_in_expr <- intersect(mutant_samples, colnames(expr_mat))
wt_samples_in_expr <- intersect(wt_samples, colnames(expr_mat))
message("Mutant samples in expression matrix: ", length(mut_samples_in_expr),
        "; WT samples in expression matrix: ", length(wt_samples_in_expr))

if (length(mut_samples_in_expr) < 2 || length(wt_samples_in_expr) < 2) {
  warning("Small group sizes in expression matrix; results may be unstable.")
}

```

## 7. Compute group means and select top 10 differential expressed genes
```{r top10_df, echo=TRUE,eval=TRUE}

# ---- compute group means and rank genes (for z-score expression we use difference-of-means) ----
mean_mut <- rowMeans(expr_mat[, mut_samples_in_expr, drop = FALSE], na.rm = TRUE)
mean_wt  <- rowMeans(expr_mat[, wt_samples_in_expr, drop = FALSE], na.rm = TRUE)
diff_mean <- mean_mut - mean_wt   # for z-scores, difference-of-means is interpretable

top_genes <- sort(diff_mean, decreasing = TRUE)
top10 <- head(top_genes, 10)
top10_df <- data.frame(gene = names(top10), diff_mean = as.numeric(top10), row.names = NULL)

print(top10_df)

```

## 8. Plot to show the top 10 differential expressed genes

```{r plot, echo=TRUE,eval=TRUE}
# ----  plot top 10 ----
ggplot(top10_df, aes(x = reorder(gene, diff_mean), y = diff_mean)) +
  geom_col() +
  coord_flip() +
  labs(title = paste0("Top 10 genes (IDH1-R132H mutant vs mean Wild type)\nStudy: ", study_id),
       x = "Gene", y = "Mean z-score difference (IDH1-R132H - WT)") +
  theme_minimal()

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
